---
title: "Boundary Logic with Credit Data: Loan-Default Prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Boundary Logic with Credit Data: Loan-Default Prediction}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)

# The SHAP comparison section needs two optional packages. Every SHAP chunk
# is gated on this flag, so the vignette still builds without them installed.
has_shap <- requireNamespace("fastshap", quietly = TRUE) &&
            requireNamespace("shapviz",  quietly = TRUE)
```

## Overview

This vignette walks through the complete three-phase Boundary Logic
workflow on a **credit-risk** problem: predicting loan default from a
dataset of 45,000 loan applications with a binary `loan_status` outcome
(1 = default, 0 = repaid). It closes with a section that places Boundary
Logic's interpretation **side by side with SHAP** for the very same
model.

The loan dataset exercises features that the simple iris example does
not:

-   **A non-linear, tree-based model** — an XGBoost classifier, wrapped
    via `bl_wrap_model()`.
-   **Domain filtering** — the credit-risk question is posed for the
    sub-population of applicants with no prior default on file.
-   **Feature pruning driven by Phase 2** — the global
    distance-to-boundary importance is used to drop weak variables and
    refit a leaner model.
-   **Business-motivated actionability constraints** — some variables
    (interest rate, credit score) move in defined directions; others
    (age) are fixed.
-   **External-applicant analysis** — scoring and explaining a brand new
    application not in the training data.

The workflow follows three phases:

-   **Phase 1:** Data preparation, model fitting, CVA biplot
    construction.
-   **Phase 2:** Global interpretation — distance to boundary,
    per-variable importance, surrogate model, and feature pruning.
-   **Phase 3:** Local interpretation of a single applicant — local
    counterfactual, Shapley attribution, sparse counterfactual.
-   **Comparison with SHAP:** global and local SHAP for the same XGBoost
    model, contrasted with the Boundary Logic output.

------------------------------------------------------------------------

## Setup

```{r load-pkg}
library(boundarylogic)
library(dplyr)
```

The final section also uses **fastshap** and **shapviz**. They are
optional (`Suggests`); if they are not installed the SHAP chunks are
skipped automatically.

------------------------------------------------------------------------

## Phase 1: Build the pipeline

### Step 1 — Load and encode

The raw dataset ships with the package in `inst/extdata`. It mixes
numeric and categorical columns, so the categorical variables are
integer-encoded first. Three high-cardinality categoricals
(`loan_intent`, `person_education`, `person_home_ownership`) are dropped
to keep the biplot legible.

```{r step1-load}
loan_raw <- read.csv(
  system.file("extdata", "loan_data.csv", package = "boundarylogic")
)

loan_encoded <- loan_raw

gender_map    <- c(female = 0, male = 1)
ownership_map <- c("MORTGAGE" = 0, "OTHER" = 1, "OWN" = 2, "RENT" = 3)
intent_map    <- c("DEBTCONSOLIDATION" = 0, "EDUCATION" = 1,
                   "HOMEIMPROVEMENT" = 2, "MEDICAL" = 3,
                   "PERSONAL" = 4, "VENTURE" = 5)
education_map <- c("High School" = 0, "Associate" = 1,
                   "Bachelor" = 2, "Master" = 3, "Doctorate" = 4)
defaults_map  <- c("No" = 0, "Yes" = 1)

loan_encoded$person_gender                  <- gender_map[loan_encoded$person_gender]
loan_encoded$person_education               <- education_map[loan_encoded$person_education]
loan_encoded$person_home_ownership          <- ownership_map[loan_encoded$person_home_ownership]
loan_encoded$loan_intent                    <- intent_map[loan_encoded$loan_intent]
loan_encoded$previous_loan_defaults_on_file <- defaults_map[loan_encoded$previous_loan_defaults_on_file]

# Coerce all columns to double so the training data and the prediction grid
# share the same type (read.csv() reads integer-valued columns as integer,
# but the grid builder generates sequences as double).
loan_encoded[] <- lapply(loan_encoded, as.numeric)

# Drop high-cardinality categoricals
vars_to_remove <- c("loan_intent", "person_education", "person_home_ownership")
loan_encoded   <- loan_encoded[, setdiff(names(loan_encoded), vars_to_remove),
                               drop = FALSE]

names(loan_encoded)
```

> **Vignette note:** to keep build time short, the steps below draw a
> seeded random subsample of the data. The full 45,000-row dataset is
> shipped in `inst/extdata/loan_data.csv` — re-run with the complete data
> for production analysis.

------------------------------------------------------------------------

### Step 2 — Exploratory biplot (no model)

A PCA biplot coloured by `loan_status` — built **before** any model — shows
the class structure and the variable loading directions. Passing no
`bl_model` to `bl_build_result()` returns a projection-only object that
`plot_biplotEZ()` renders in the usual style.

```{r step2-explore}
set.seed(121L)
loan_explore <- loan_encoded[sample(nrow(loan_encoded), 4000L), ]

bl_dat_exp <- bl_prepare_data(
  data           = loan_explore,
  class_col      = "loan_status",
  train_fraction = 1,
  seed           = 121L,
  hull_fraction  = 1
)

bl_proj <- bl_build_result(
  bl_data = bl_dat_exp,
  method  = "PCA",
  title   = "Loan default - exploratory PCA biplot"
)

plot_biplotEZ(
  bl_proj,
  label_dir         = "Paral",
  label_offset_var  = c("person_age", "person_gender", "person_income",
                        "person_emp_exp", "loan_amnt", "loan_int_rate",
                        "loan_percent_income", "cb_person_cred_hist_length",
                        "credit_score", "previous_loan_defaults_on_file"),
  label_offset_dist = c(0, 0, 0.5, 1, 0.5, 1, 0, 1.5, 0, 0)
)
```

------------------------------------------------------------------------

### Step 3 — Domain filter

The credit-risk model targets applicants with **no prior default on
file**. After filtering, `previous_loan_defaults_on_file` is constant and
is excluded from the feature set (zero variance). The modelling
population is then subsampled to 4,000 rows for a fast build.

```{r step3-filter}
loan_filtered <- loan_encoded[loan_encoded$previous_loan_defaults_on_file == 0, ]

set.seed(121L)
loan_filtered <- loan_filtered[sample(nrow(loan_filtered), 4000L), ]

feature_cols <- setdiff(
  names(loan_filtered),
  c("loan_status", "previous_loan_defaults_on_file")
)

cat("Rows:", nrow(loan_filtered), "\n")
cat("Features:", paste(feature_cols, collapse = ", "), "\n")
```

------------------------------------------------------------------------

### Steps 4-6 — Prepare data, fit XGBoost, build the biplot

`bl_prepare_data()` makes a reproducible 80/20 split, encodes the class,
and trims the outermost 10% of training points (`hull_fraction = 0.9`).
An XGBoost classifier is then fitted directly and registered with
`bl_wrap_model()` using the built-in `"XGB"` dispatch — the model must be
supplied as `list(model = <booster>, features = <character vector>)`.

```{r steps4-6}
bl_dat <- bl_prepare_data(
  data           = loan_filtered,
  class_col      = "loan_status",
  feature_cols   = feature_cols,
  train_fraction = 0.8,
  seed           = 121L,
  hull_fraction  = 0.9
)

xgb_data <- xgboost::xgb.DMatrix(
  data  = as.matrix(bl_dat$train_data[, bl_dat$var_names]),
  label = bl_dat$train_data$class
)
xgb_fit <- xgboost::xgb.train(
  params  = list(objective     = "binary:logistic",
                 eval_metric   = "logloss",
                 max_depth     = 3,
                 learning_rate = 0.1),
  data    = xgb_data,
  nrounds = 200,
  verbose = 0
)

bl_mod <- bl_wrap_model(
  model      = list(model = xgb_fit, features = bl_dat$var_names),
  model_type = "XGB",
  var_names  = bl_dat$var_names,
  train_data = bl_dat$train_data
)

bl_results <- bl_build_result(
  bl_data  = bl_dat,
  bl_model = bl_mod,
  method   = "CVA",
  title    = "Loan default (no prior default) - XGB, CVA biplot",
  b_margin = 0.001
)

print(bl_results)
```

**CVA projection** is the default for this workflow: it maximises
between-class separation in the 2D plane, making the decision boundary
clearly visible.

```{r biplot-train}
plot_biplotEZ(
  bl_results,
  label_dir         = "Hor",
  label_offset_var  = 0L,
  label_offset_dist = 0.5
)
```

Overlay the test observations (`filter_to_polygon = TRUE` drops points
that project outside the training hull):

```{r biplot-test}
test_pts <- bl_project_points(bl_results$test_data, bl_results,
                              filter_to_polygon = TRUE)
plot(bl_results, points = test_pts)
```

------------------------------------------------------------------------

## Phase 2: Global interpretation

### Step 7 — Find nearest boundary point: `bl_find_boundary()`

`bl_find_boundary()` finds, for every observation, the nearest point on
the decision boundary and back-projects it to feature space. The search
is constrained to the training feature envelope (`train_ranges`).

```{r step7-boundary}
bl_bnd <- bl_find_boundary(bl_results)
print(bl_bnd)

plot_biplotEZ(bl_results, points = test_pts)
```

```{r step7-pick, eval=FALSE}
# Interactive: click an observation to see its counterfactual arrow
bl_pick_point(bl_results, bl_boundary = bl_bnd)
```

------------------------------------------------------------------------

### Step 8 — Distance-to-boundary plot

`plot(bl_bnd)` decomposes each observation's distance to the boundary
into a signed, standardised contribution per variable, sorted from least
to most important.

-   **Y-axis:** `variable : total absolute standardised distance`.
-   **X-axis:** signed standardised distance.
-   **Colours:** TP = red, TN = blue, FP = purple, FN = orange.

```{r step8-jitter}
plot(bl_bnd)
```

```{r step8-boxplot}
plot(bl_bnd, type = "boxplot")
```

------------------------------------------------------------------------

### Step 9 — Surrogate model: `bl_surrogate()`

`bl_surrogate()` classifies each training observation purely from its
position relative to the surrogate contour lines in Z-space, giving a
spatial approximation of the classifier confined to the training hull.

```{r step9-surrogate}
bl_surr <- bl_surrogate(bl_results)
print(bl_surr)
plot(bl_surr)
```

------------------------------------------------------------------------

### Step 10 — Prune weak variables and refit

The per-variable totals from `bl_find_boundary()` act as a global
importance ranking. `bl_robustness()` exposes them programmatically as
`sum_of_distance`. The weakest variables are dropped and the model is
refit on the leaner feature set.

```{r step10-prune}
rob     <- bl_robustness(bl_bnd)
var_imp <- sort(rob$sum_of_distance)        # ascending: weakest first
print(round(var_imp, 2))

# Drop the four weakest variables; keep the five strongest.
vars_to_drop    <- c("person_gender", "person_emp_exp",
                     "cb_person_cred_hist_length", "person_income")
feature_cols_v2 <- setdiff(bl_dat$var_names, vars_to_drop)
cat("Retained features:", paste(feature_cols_v2, collapse = ", "), "\n")
```

Refit the whole pipeline on the reduced feature set:

```{r step10-refit}
bl_dat_v2 <- bl_prepare_data(
  data           = loan_filtered,
  class_col      = "loan_status",
  feature_cols   = feature_cols_v2,
  train_fraction = 0.8,
  seed           = 121L,
  hull_fraction  = 0.9
)

xgb_data_v2 <- xgboost::xgb.DMatrix(
  data  = as.matrix(bl_dat_v2$train_data[, bl_dat_v2$var_names]),
  label = bl_dat_v2$train_data$class
)
xgb_fit_v2 <- xgboost::xgb.train(
  params  = list(objective     = "binary:logistic",
                 eval_metric   = "logloss",
                 max_depth     = 3,
                 learning_rate = 0.1),
  data    = xgb_data_v2,
  nrounds = 200,
  verbose = 0
)
bl_mod_v2 <- bl_wrap_model(
  model      = list(model = xgb_fit_v2, features = bl_dat_v2$var_names),
  model_type = "XGB",
  var_names  = bl_dat_v2$var_names,
  train_data = bl_dat_v2$train_data
)

bl_results_v2 <- bl_build_result(
  bl_data  = bl_dat_v2,
  bl_model = bl_mod_v2,
  method   = "CVA",
  title    = "Loan default - XGB, reduced features, CVA biplot",
  b_margin = 0.01
)

plot_biplotEZ(bl_results_v2)
test_pts_v2 <- bl_project_points(bl_results_v2$test_data, bl_results_v2)
plot(bl_results_v2, points = test_pts_v2)
```

Re-inspect the reduced model globally:

```{r step10-refit-global}
bl_bnd_v2 <- bl_find_boundary(bl_results_v2)
plot(bl_bnd_v2)

bl_surr_v2 <- bl_surrogate(bl_results_v2)
plot(bl_surr_v2,
     label_dir         = "Paral",
     label_offset_var  = c("person_age", "loan_amnt", "loan_int_rate",
                           "loan_percent_income", "credit_score"),
     label_offset_dist = c(0, 0, 0.5, 0, 0),
     ticks_v = 5)
```

------------------------------------------------------------------------

## Phase 3: Local interpretation

All of Phase 3 uses the reduced model `bl_results_v2`.

### Step 11 — Inspect predictions and select a target

`bl_predict()` scores every test observation. **False Negatives** (true
class 1, predicted 0) are the most actionable: the applicant will default
but the model clears them — what would have to change to flag them
correctly?

```{r step11-select}
pred_summary <- bl_predict(bl_results_v2)

# Pick the first false-negative applicant; fall back to the first true
# defaulter so the vignette knits regardless of the data split.
fn_idx <- which(pred_summary$true_class == 1 & pred_summary$pred_class == 0)
tdp    <- if (length(fn_idx) > 0) fn_idx[1] else which(pred_summary$true_class == 1)[1]
pred_summary[tdp, ]

tgt <- bl_select_target(bl_results_v2, target = tdp)
print(tgt)

plot(
  bl_results_v2,
  points       = test_pts_v2,
  target_point = bl_results_v2$test_data[tdp, ],
  target_label = tdp,
  label_dir         = "Paral",
  label_offset_var  = c("person_age", "loan_amnt", "loan_int_rate",
                        "loan_percent_income", "credit_score"),
  label_offset_dist = c(0, 0, 0.5, 0, 0),
  ticks_v = 3
)
```

------------------------------------------------------------------------

### Step 12 — Set actionability constraints: `set_filters()`

Not every variable can move freely in a lending decision. `set_filters()`
encodes the business constraints:

-   **`"decrease"`** — counterfactual value must be ≤ observed.
-   **`"increase"`** — counterfactual value must be ≥ observed.
-   **`"fixed"`** — not actionable; the search allows ±0.5 of observed and
    the sparse counterfactual always reverts it.
-   **`c(min, max)`** — counterfactual must lie within this absolute
    range.

Here `person_age` is fixed (not actionable), while `loan_int_rate` and
`credit_score` are constrained to increase. Only variables still in the
reduced model may be named.

```{r step12-filters}
flt <- set_filters(
  tgt,
  person_age    = "fixed",      # not actionable
  loan_int_rate = "increase",   # set by the lender, not the applicant
  credit_score  = "increase"
)
print(flt)
```

------------------------------------------------------------------------

### Step 13 — Find local counterfactual: `bl_find_local_cf()`

`bl_find_local_cf()` rotates the biplot around the target, builds a local
prediction grid, applies `train_ranges` and the actionability filters,
and finds the nearest opposing-class boundary point — repeating across
eigenvector pairs and keeping the closest.

```{r step13-local-cf}
bl_local <- bl_find_local_cf(
  bl_result   = bl_results_v2,
  bl_target   = tgt,
  set_filters = flt
)
print(bl_local)
```

------------------------------------------------------------------------

### Step 14 — Local biplot

`plot(bl_local)` renders the rotated eigenvector plane that minimises the
distance to the boundary, showing the target (filled circle) and the
counterfactual (grey cross with arrow).

```{r step14-local-biplot}
plot(bl_local,
     label_dir         = "Paral",
     label_offset_var  = c("person_age", "loan_amnt", "loan_int_rate",
                           "loan_percent_income", "credit_score"),
     label_offset_dist = c(0, 0.5, 0, 0, 0),
     ticks_var = c("person_age", "loan_amnt", "loan_int_rate",
                   "loan_percent_income", "credit_score"),
     ticks_n = c(2, 4, 200, 200, 20))
```

------------------------------------------------------------------------

### Step 15 — Shapley attribution: `bl_shapley()`

`bl_shapley()` decomposes the path from the target to the counterfactual
into per-variable contributions. Each variable is classified as
**Supports** (helps cross the boundary) or **Contradicts** (works against
it). Y-axis labels read `variable: observed_value -> required_change`.

```{r step15-shapley}
bl_shapley_values <- bl_shapley(bl_local)
print(bl_shapley_values)
plot(bl_shapley_values)
```

------------------------------------------------------------------------

### Step 16 — Sparse counterfactual: `bl_find_sparse_cf()`

`bl_find_sparse_cf()` reverts every **Contradicts** variable (and every
`"fixed"` variable) to its observed value, so only the **Supports**
variables change. `solution_valid = TRUE` confirms the sparse change
still flips the prediction.

```{r step16-sparse}
bl_sparse <- bl_find_sparse_cf(bl_shapley_values, round_to = NULL)
print(bl_sparse)
plot(bl_sparse)
```

------------------------------------------------------------------------

## Alternative: unconstrained local search

Running `bl_find_local_cf()` without actionability constraints gives the
geometrically nearest boundary point — a useful baseline.

```{r step17-unconstrained}
bl_local_free  <- bl_find_local_cf(bl_results_v2, tgt)
plot(bl_local_free,
     label_dir         = "Paral",
     label_offset_var  = c("person_age", "loan_amnt", "loan_int_rate",
                           "loan_percent_income", "credit_score"),
     label_offset_dist = c(0, 0, 0, 0, 1),
     ticks_var = c("person_age", "loan_amnt", "loan_int_rate",
                   "loan_percent_income", "credit_score"),
     ticks_n = c(2, 4, 200, 200, 20))

bl_shap_free   <- bl_shapley(bl_local_free)
bl_sparse_free <- bl_find_sparse_cf(bl_shap_free, round_to = NULL)
plot(bl_shap_free)
plot(bl_sparse_free)
```

------------------------------------------------------------------------

## External applicant

Phase 3 is not limited to the train/test set. Pass a single-row data
frame as the `target` to score and explain a new application. Only the
retained features are needed.

```{r step18-external}
new_applicant <- data.frame(
  person_age          = 32,
  loan_amnt           = 12000,
  loan_int_rate       = 13.5,
  loan_percent_income = 12000 / 248000,
  credit_score        = 420
)
new_applicant <- new_applicant[, intersect(names(new_applicant), feature_cols_v2),
                               drop = FALSE]

plot_biplotEZ(
  bl_results_v2,
  points       = test_pts_v2,
  target_point = new_applicant,
  target_label = "new"
)

tgt_ext       <- bl_select_target(bl_results_v2, target = new_applicant)
bl_local_ext  <- bl_find_local_cf(bl_results_v2, tgt_ext)
bl_shap_ext   <- bl_shapley(bl_local_ext)
bl_sparse_ext <- bl_find_sparse_cf(bl_shap_ext)

plot(bl_local_ext)
plot(bl_shap_ext)
plot(bl_sparse_ext)
```

------------------------------------------------------------------------

## Comparison with SHAP

SHAP (SHapley Additive exPlanations) is the most widely used post-hoc
attribution method, so it is the natural yardstick for Boundary Logic.
Both rest on Shapley values, but they answer **different questions**:

-   **SHAP** attributes the *model's predicted probability* for an
    observation, relative to the average (baseline) prediction over the
    training data. It says *why the score is what it is*.
-   **Boundary Logic** attributes *distance to the decision boundary*
    (globally) and *the minimal change that flips the prediction*
    (locally). It says *what would have to change to alter the decision*.

The chunks below fit real SHAP on the **same XGBoost models** using
`fastshap` and `shapviz`. Each chunk is guarded with `if (has_shap)`, so
the section renders only when both packages are installed (and the
vignette still builds when they are not).

```{r shap-missing, echo = FALSE, results = "asis"}
if (!has_shap) {
  cat("> **SHAP packages not available.** Install `fastshap` and `shapviz`",
      "to render this section:\n>\n>",
      "`install.packages(c(\"fastshap\", \"shapviz\"))`\n")
}
```

```{r shap-setup}
if (has_shap) {
  # fastshap needs a prediction wrapper f(object, newdata) -> numeric vector.
  # For XGBoost the new data must be wrapped in an xgb.DMatrix.
  pred_fn <- function(object, newdata) {
    as.numeric(predict(object, xgboost::xgb.DMatrix(as.matrix(newdata))))
  }

  # Background / explanation data for each model
  X_train_full <- bl_dat$train_data[,    bl_dat$var_names,    drop = FALSE]
  X_train_v2   <- bl_dat_v2$train_data[, bl_dat_v2$var_names, drop = FALSE]
  X_test_v2    <- bl_dat_v2$test_data[,  bl_dat_v2$var_names, drop = FALSE]

  # SHAP baseline = mean predicted probability over the (reduced-model)
  # training background. fastshap does not attach one, so shapviz needs it.
  shap_baseline <- mean(pred_fn(xgb_fit_v2, X_train_v2))
}
```

### Global: SHAP vs. Boundary Logic variable selection

**(a) Which five variables would SHAP have kept?**

In Phase 2 the *distance-to-boundary* importance picked the five
variables in `feature_cols_v2`. Here we ask SHAP the same question on the
**full nine-feature** model: rank the variables by mean absolute SHAP
value and take its top five.

```{r shap-global-select}
if (has_shap) {
  set.seed(42L)
  shap_full <- fastshap::explain(
    object       = xgb_fit,            # full-feature model
    X            = X_train_full,
    pred_wrapper = pred_fn,
    nsim         = 50L
  )

  shap_imp  <- sort(colMeans(abs(as.matrix(shap_full))), decreasing = TRUE)
  shap_top5 <- names(shap_imp)[1:5]

  # Distance-to-boundary ranking (strongest first) from Phase 2
  dist_top5 <- names(sort(rob$sum_of_distance, decreasing = TRUE))[1:5]

  print(data.frame(
    rank          = 1:5,
    SHAP          = shap_top5,
    dist_to_bound = dist_top5
  ))
}
```

The two criteria usually agree on the dominant credit-risk drivers
(interest rate, loan-to-income, credit score) but can disagree in the
tail: SHAP ranks a variable by how much it moves the *predicted
probability* away from the average applicant, whereas Boundary Logic
ranks it by its total standardised contribution to *distance from the
decision boundary*. A variable can shift many scores a little (high SHAP)
yet rarely be the lever that carries a point across the boundary (low
distance contribution), and vice versa.

**(b) Global SHAP of the retained five variables**

Now run SHAP on the reduced model — exactly the five variables the
workflow proceeded with — and view the standard global summaries.

```{r shap-global-plots}
if (has_shap) {
  set.seed(42L)
  shap_v2 <- fastshap::explain(
    object       = xgb_fit_v2,         # reduced model
    X            = X_train_v2,
    pred_wrapper = pred_fn,
    nsim         = 50L
  )
  sv_v2 <- shapviz::shapviz(as.matrix(shap_v2), X = X_train_v2,
                            baseline = shap_baseline)

  print(shapviz::sv_importance(sv_v2, kind = "bar") +
          ggplot2::labs(title = "Global SHAP importance - reduced XGB model"))
}
```

```{r shap-global-beeswarm}
if (has_shap) {
  print(shapviz::sv_importance(sv_v2, kind = "beeswarm") +
          ggplot2::labs(title = "Global SHAP beeswarm - reduced XGB model"))
}
```

Compare these against `plot(bl_bnd_v2)` (Step 10) and `bl_surrogate()`
(Step 9). SHAP's bar chart and the Boundary Logic distance ranking are
both global importance measures, but the SHAP beeswarm additionally shows
the *direction* in which high/low feature values push the predicted
probability, while the Boundary Logic plot shows the *signed distance*
each observation sits from the boundary along each variable.

### Local: SHAP vs. `bl_shapley` attribution

For the **same target applicant** explained in Phase 3, compute local
SHAP and view the waterfall and force plots.

```{r shap-local}
if (has_shap) {
  set.seed(42L)
  shap_local <- fastshap::explain(
    object       = xgb_fit_v2,
    X            = X_train_v2,
    pred_wrapper = pred_fn,
    nsim         = 200L,
    newdata      = X_test_v2[tdp, , drop = FALSE],
    adjust       = TRUE          # force additivity: sum(shap) = f(x) - baseline
  )
  sv_local <- shapviz::shapviz(as.matrix(shap_local),
                               X = X_test_v2[tdp, , drop = FALSE],
                               baseline = shap_baseline)

  print(shapviz::sv_waterfall(sv_local) +
          ggplot2::labs(title = sprintf("Local SHAP waterfall - applicant %d", tdp)))
}
```

```{r shap-local-force}
if (has_shap) {
  print(shapviz::sv_force(sv_local) +
          ggplot2::labs(title = sprintf("Local SHAP force plot - applicant %d", tdp)))
}
```

Put next to `plot(bl_shapley_values)` (Step 15), the contrast is sharp:

| Aspect | Local SHAP (waterfall/force) | `bl_shapley()` |
|---|---|---|
| What it explains | The predicted probability for this applicant | The path from this applicant to the boundary counterfactual |
| Baseline | Mean training prediction | The applicant's observed feature values |
| Reading | Each bar pushes the score up/down | Each bar **Supports** or **Contradicts** crossing the boundary |
| Output | A number (the score decomposition) | An actionable change (`observed -> required`) |
| Constraints | None | Honours `set_filters()` actionability |

SHAP tells you **why the model scored this applicant as it did**;
`bl_shapley()` tells you **the minimal, constraint-respecting change that
would flip the decision**. They are complementary: SHAP for understanding
the score, Boundary Logic for acting on it.

------------------------------------------------------------------------

## Summary

| Phase | Step | Function | Purpose |
|-------|------|----------|---------|
| 1 | 1-3 | `bl_prepare_data()` | Load, encode, domain-filter, split, hull-trim |
| 1 | 2 | `bl_build_result(method="PCA")` | Exploratory biplot (no model) |
| 1 | 4-6 | `bl_wrap_model()` + `bl_build_result()` | XGBoost + CVA biplot |
| 2 | 7 | `bl_find_boundary()` | Global counterfactual search |
| 2 | 8 | `plot(bl_bnd)` | Distance-to-boundary importance |
| 2 | 9 | `bl_surrogate()` | Spatial surrogate model |
| 2 | 10 | `bl_robustness()` | Prune weak variables, refit |
| 3 | 11 | `bl_predict()` + `bl_select_target()` | Select target applicant |
| 3 | 12 | `set_filters()` | Encode actionability constraints |
| 3 | 13 | `bl_find_local_cf()` | Nearest constrained counterfactual |
| 3 | 14 | `plot(bl_local)` | Local biplot |
| 3 | 15 | `bl_shapley()` | Shapley attribution of the flip |
| 3 | 16 | `bl_find_sparse_cf()` | Sparse counterfactual |
| - | - | `fastshap` / `shapviz` | SHAP comparison (global + local) |

Boundary Logic and SHAP are best used together. SHAP explains the
model's score relative to an average applicant; Boundary Logic locates
each applicant relative to the decision boundary and identifies the
smallest, business-feasible change that would alter the lending decision.
