---
title: "Boundary Logic with Real Clinical Data: The Pima Diabetes Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Boundary Logic with Real Clinical Data: The Pima Diabetes Dataset}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette walks through the complete three-phase Boundary Logic
workflow using the **Pima Indians Diabetes** dataset — a real-world
clinical dataset with 768 observations, 8 numeric predictors, and a
binary outcome (diabetic / non-diabetic).

The Pima dataset demonstrates features that do not arise in the simple
iris example:

-   **Non-linear decision surface** — handled here with a Generalised
    Additive Model (GAM) using spline terms, wrapped via
    `bl_wrap_model()`.
-   **Clinically motivated actionability constraints** — some variables
    (Glucose) can be influenced by treatment, while others (Age,
    Pregnancies) cannot.
-   **External patient analysis** — predicting and explaining the
    outcome for a new patient not in the training set.

The workflow follows three phases:

-   **Phase 1:** Data preparation, model fitting, CVA biplot
    construction.
-   **Phase 2:** Global interpretation — distance to boundary,
    per-variable importance, surrogate model.
-   **Phase 3:** Local interpretation of a single patient — local
    counterfactual, Shapley attribution, sparse counterfactual.

------------------------------------------------------------------------

## Setup

```{r load-pkg}
library(boundarylogic)
library(dplyr)
```

------------------------------------------------------------------------

## Phase 1: Build the pipeline

### Step 1 — Prepare data

The dataset has 8 numeric features and a binary `Outcome` column (1 =
diabetic, 0 = non-diabetic). Before fitting the model, one data quality
step is applied:

-   **Filter implausible zero values** — `Insulin` and `SkinThickness`
    cannot physiologically be zero; those rows are removed.

`bl_prepare_data()` converts the `Outcome` column to `class` (it is
already 0/1 so `target_class = NULL`) and creates a reproducible 80/20
train/test split.

```{r step1}
pima_csv <- system.file("extdata", "pima_diabetes.csv",
                        package = "boundarylogic")

pima_raw <- read.csv(pima_csv)

pima_raw <- pima_raw %>%
  filter(Insulin > 0, SkinThickness > 0)

bl_dat <- bl_prepare_data(
  data           = pima_raw,
  class_col      = "Outcome",
  target_class   = NULL,       # already 0/1
  train_fraction = 0.8,
  seed           = 121L
)

print(bl_dat)
```

**What to check:**

-   `Features` — the eight clinical measurement columns.
-   `Train rows / Test rows` — approximately 80/20 of the filtered
    dataset.
-   `Class balance` — check that both classes are represented in the
    training set.

```{r step1-check}
table(bl_dat$train_data$class)
```

------------------------------------------------------------------------

### Step 2 — Filter outliers (optional)

`bl_filter_outliers()` removes training observations that fall outside
the convex hull enclosing the specified fraction of the data. With
`hull_fraction = 0.9`, the outermost 10 % of training points are removed
before model fitting.

```{r step2}
bl_filt <- bl_filter_outliers(bl_dat, hull_fraction = 0.9)
print(bl_filt)
```

------------------------------------------------------------------------

### Step 3 — Fit a GAM and wrap it

The Pima outcome is non-linear with respect to most predictors. A
Generalised Additive Model (GAM) with spline terms is a good choice
here: it is flexible enough to capture the non-linear decision surface
while remaining interpretable.

`mgcv::gam()` is fitted directly (giving full control over the formula
and spline settings). The fitted object is then registered as a
`bl_model` via `bl_wrap_model()` with a custom `predict_fn`.

```{r step3}
gam_formula <- class ~ s(Glucose) + s(BloodPressure) +
  s(Pregnancies) + s(SkinThickness) + Insulin +
  s(BMI) + DiabetesPedigreeFunction + Age

custom_gam <- mgcv::gam(
  formula = gam_formula,
  data    = bl_filt$train_data,
  family  = binomial(link = "logit")
)

bl_mod <- bl_wrap_model(
  model      = custom_gam,
  model_type = "custom",
  var_names  = bl_filt$var_names,
  predict_fn = function(m, new_data) {
    as.numeric(mgcv::predict.gam(m, newdata = new_data, type = "response"))
  },
  train_data = bl_filt$train_data
)

print(bl_mod)
```

**Why `model_type = "custom"`?**\
`mgcv::gam()` is not one of the built-in parsnip-backed types. Supplying
`predict_fn` tells the package exactly how to call `predict()` on the
model object. The function signature must be `function(model, new_data)`
returning a numeric probability vector.

------------------------------------------------------------------------

### Steps 4–6 — Build and assemble

`bl_build_result()` combines the projection (Step 4), prediction grid
(Step 5), and assembly (Step 6) into a single call.

**CVA projection** is used here. CVA (Canonical Variate Analysis)
maximises between-class separation relative to within-class spread,
which makes the decision boundary clearly visible in the 2D biplot plane
for datasets with a well-separated binary outcome.

`rounding = 2L` sets a wider contour band (`cutoff ± 0.01`, i.e. contour
levels at 0.49 / 0.51), which works well for smooth, gradual decision
surfaces like those produced by a GAM:

| `rounding` | `b_margin` | Contour levels    |
|------------|------------|-------------------|
| `2L`       | `0.01`     | `0.49` / `0.51`   |
| `3L`       | `0.001`    | `0.499` / `0.501` |

```{r steps4-6}
bl_results <- bl_build_result(
  bl_data  = bl_filt,
  bl_model = bl_mod,
  method   = "CVA",
  title    = "Pima diabetes — GAM, CVA biplot",
  rounding = 2L
)

print(bl_results)
```

```{r steps4-6-summary}
summary(bl_results)
```

------------------------------------------------------------------------

### Visualise: `plot_biplotEZ()`

The biplot shows the CVA prediction surface (blue = non-diabetic, red =
diabetic probability regions), training observations coloured by
confusion category, and the decision boundary contour:

```{r biplot-train}
plot_biplotEZ(bl_results)
```

Overlay the test observations:

```{r biplot-test}
test_pts <- bl_project_points(bl_results$test_data, bl_results)
plot_biplotEZ(bl_results, points = test_pts)
```

------------------------------------------------------------------------

## Phase 2: Global interpretation

Phase 2 uses the `bl_results` object to interpret the model at the
population level: finding the nearest boundary point for every
observation, measuring variable-level distances, and fitting a surrogate
model.

------------------------------------------------------------------------

### Step 7 — Find nearest boundary point: `bl_find_boundary()`

`bl_find_boundary()` searches the decision boundary contour for each
observation and back-projects the nearest boundary point to X-space. The
result summarises how far each observation is from the decision boundary
and which variables drive that distance.

The search is constrained to `train_ranges` — counterfactuals must lie
within the range of the training feature envelope.

```{r step7-boundary}
bl_bnd <- bl_find_boundary(bl_results)
print(bl_bnd)
```

Overlay the boundary arrows on the biplot (each arrow runs from an
observation to its nearest boundary point):

```{r step7-biplot}
plot_biplotEZ(bl_results, points = test_pts, boundary = bl_bnd)
```

------------------------------------------------------------------------

### Step 8 — Distance-to-boundary plot

`plot(bl_bnd)` decomposes each observation's distance to the boundary
into a signed, standardised contribution per variable. Variables are
sorted from least to most important (bottom to top).

-   **Y-axis:** `variable : total absolute standardised distance`
    (importance proxy).
-   **X-axis:** signed standardised distance — positive means the
    observation is above the boundary in that variable's direction.
-   **Colours:** TP = red, TN = blue, FP = purple, FN = orange.
-   The overall **robustness scalar** and **per-variable totals** are
    printed to the console.

```{r step8-jitter}
plot(bl_bnd)
```

```{r step8-boxplot}
plot(bl_bnd, type = "boxplot")
```

> **Note:** `plot(bl_bnd)` already prints the overall robustness scalar
> and per-variable totals to the console. Call `bl_robustness(bl_bnd)`
> separately only when you need to access the values programmatically —
> for example, to compare robustness across models.

------------------------------------------------------------------------

### Step 9 — Surrogate model: `bl_surrogate()`

`bl_surrogate()` assigns each training observation a class based solely
on its position in Z-space relative to the hull-clipped surrogate
contour lines. This provides a purely spatial approximation of the
classifier confined to the training hull. Observations that project
outside the training hull receive `surrogate_pred = NA`.

Two accuracy measures are reported:

-   `accuracy_vs_model` — agreement with the fitted GAM's predictions.
-   `accuracy_vs_labels` — accuracy against true class labels.

```{r step9-surrogate}
bl_surr <- bl_surrogate(bl_results)
print(bl_surr)
plot(bl_surr)
```

------------------------------------------------------------------------

## Phase 3: Local interpretation

Phase 3 explains a single patient observation by finding the nearest
counterfactual via biplot rotation, decomposing the required changes
using Shapley values, and producing a sparse counterfactual.

------------------------------------------------------------------------

### Step 10 — Inspect predictions and select a target

`bl_predict()` scores every test observation and returns a tidy data
frame. Inspect it to choose the patient you want to explain, then wrap
that row in a `bl_target` object using `bl_select_target()`.

```{r step10-select}
pred_summary <- bl_predict(bl_results)
print(pred_summary)

# Select the first predicted diabetic patient (class 1) for local explanation.
# With seed = 121L this is row 1 of test_data; the dynamic selection below
# ensures the vignette knits correctly regardless of the data split.
tdp <- which(pred_summary$pred_class == 1)[1]
tgt <- bl_select_target(bl_results, target = tdp)
print(tgt)

# Highlight the target on the main biplot
plot_biplotEZ(
  bl_results,
  points       = test_pts,
  target_point = unlist(tgt$x_obs),
  target_label = tdp
)
```

------------------------------------------------------------------------

### Step 12 — Set actionability constraints: `set_filters()`

In a clinical context, not all variables can be changed for a given
patient. `set_filters()` encodes these constraints:

-   **`"decrease"`** — the counterfactual value must be ≤ observed
    (e.g., lowering Glucose via diet/medication is a realistic
    intervention).
-   **`"fixed"`** — the variable is not actionable; the search allows
    ±0.5 of the observed value, and the sparse counterfactual always
    reverts to the original observed value.
-   **`"increase"`** — the counterfactual value must be ≥ observed.
-   **`c(min, max)`** — counterfactual must lie within this absolute
    range.

For this patient, `Glucose` is the primary clinical lever (can be
reduced), while `Age` and `Pregnancies` are biological facts that cannot
be changed:

```{r step12-filters}
flt <- set_filters(
  tgt,
  Glucose     = "decrease",   # reducing Glucose is the clinical lever
  Age         = "fixed",      # Age is not actionable
  Pregnancies = "fixed"       # Pregnancies is not actionable
)

print(flt)
```

------------------------------------------------------------------------

### Step 13 — Find local counterfactual: `bl_find_local_cf()`

`bl_find_local_cf()` searches for the nearest decision boundary point
relative to the target patient by testing multiple eigenvector pairs.
For each pair it:

1.  Rotates the biplot so the target lies on the projection plane.
2.  Builds a local prediction grid in the rotated Z-space.
3.  Applies `train_ranges` (feasibility) and the actionability filters
    from Step 12.
4.  Finds the nearest opposing-class boundary point.

The pair with the overall minimum Z-space distance is selected.

> **Note:** The Z-space convex hull polygon is not applied during the
> local search — rotation invalidates the original polygon coordinates.
> Only `train_ranges` and `set_filters` constrain the search.

```{r step13-local-cf}
bl_local <- bl_find_local_cf(
  bl_result   = bl_results,
  bl_target   = tgt,
  set_filters = flt
)

print(bl_local)
```

------------------------------------------------------------------------

### Step 14 — Local biplot

`plot(bl_local)` renders the biplot in the rotated eigenvector plane
that minimises the distance to the boundary. By default, training
observations are hidden — only the target (filled circle) and
counterfactual (grey cross with arrow) are shown.

```{r step14-local-biplot}
plot(bl_local)
```

To also show training observations:

```{r step14-with-training}
plot(bl_local, no_points = FALSE)
```

------------------------------------------------------------------------

### Step 15 — Shapley attribution: `bl_shapley()`

`bl_shapley()` decomposes the path from the target to the counterfactual
into per-variable contributions using the exact Shapley formula. Each
variable is classified as:

-   **Supports** (dark blue) — moving in this direction helps cross the
    boundary (reduces the diabetic probability toward 0.5).
-   **Contradicts** (grey) — moving in this direction works against
    crossing the boundary.

Y-axis labels show `variable: observed_value → required_change`, both
rounded to 3 decimal places.

```{r step15-shapley}
bl_shapley_values <- bl_shapley(bl_local)
print(bl_shapley_values)

plot(bl_shapley_values)
```

------------------------------------------------------------------------

### Step 16 — Sparse counterfactual: `bl_find_sparse_cf()`

`bl_find_sparse_cf()` creates a parsimonious counterfactual by reverting
all **Contradicts** variables to their observed values, and reverting
all variables marked `"fixed"` in `set_filters()`. Only the **Supports**
variables take their counterfactual values.

`solution_valid = TRUE` confirms the sparse counterfactual still crosses
the decision boundary (predicted probability changes from class 1 to
class 0).

```{r step16-sparse}
bl_sparse <- bl_find_sparse_cf(bl_shapley_values, round_to = NULL)

# print() shows three predictions to verify the class flip:
#   Observed  — model score for the original patient
#   Full CF   — model score at the full counterfactual
#   Sparse CF — model score for the sparse (zeroed) counterfactual
print(bl_sparse)

# Grey cross   = full counterfactual (all variables changed)
# Green cross  = sparse CF (valid — crosses boundary)
# Yellow cross = sparse CF (invalid — does not cross boundary)
plot(bl_sparse)
```

------------------------------------------------------------------------

## Step 17 (Alternative): Unconstrained local search

Running `bl_find_local_cf()` without actionability constraints is useful
as a baseline — it shows the geometrically nearest boundary point
regardless of clinical plausibility. If the constrained search in Step
13 returned `solution_found = FALSE`, the unconstrained search confirms
whether a boundary point exists at all.

```{r step17-unconstrained}
bl_local_free   <- bl_find_local_cf(bl_results, tgt)
print(bl_local_free)

plot(bl_local_free)

bl_shap_free   <- bl_shapley(bl_local_free)
bl_sparse_free <- bl_find_sparse_cf(bl_shap_free, round_to = NULL)

plot(bl_shap_free)
plot(bl_sparse_free)
```

------------------------------------------------------------------------

## Step 18: External patient

Phase 3 is not limited to observations in the train/test set. Pass a
single-row data frame as the `target` argument to `bl_select_target()`
to analyse a new patient who was not part of the original data.

The feature values below correspond to Patient 1 from the original Pima
dataset (before the Insulin/SkinThickness filter was applied):

```{r step18-external}
new_patient <- data.frame(
  Pregnancies              = 2,
  Glucose                  = 148,
  BloodPressure            = 72,
  SkinThickness            = 35,
  Insulin                  = 100,
  BMI                      = 33.6,
  DiabetesPedigreeFunction = 0.627,
  Age                      = 50
)

tgt_ext <- bl_select_target(bl_results, target = new_patient)
print(tgt_ext)

# Highlight on the main biplot
plot_biplotEZ(
  bl_results,
  points       = test_pts,
  target_point = unlist(tgt_ext$x_obs),
  target_label = "new"
)
```

Run the full local explanation pipeline for the external patient:

```{r step18-local}
bl_local_ext <- bl_find_local_cf(bl_results, tgt_ext)
print(bl_local_ext)

plot(bl_local_ext)
```

```{r step18-shapley}
bl_shap_ext   <- bl_shapley(bl_local_ext)
bl_sparse_ext <- bl_find_sparse_cf(bl_shap_ext)

plot(bl_shap_ext)
plot(bl_sparse_ext)
```

------------------------------------------------------------------------

## Summary

This vignette demonstrated the complete Boundary Logic workflow on a
real clinical dataset:

| Phase | Step | Function                              | Purpose                                   |
|-------|------|---------------------------------------|-------------------------------------------|
| 1     | 1    | `bl_prepare_data()`                   | Load and split the Pima dataset           |
| 1     | 2    | `bl_filter_outliers()`                | Remove the outer 10 % of training points  |
| 1     | 3    | `bl_wrap_model()`                     | Register a custom GAM with spline terms   |
| 1     | 4–6  | `bl_build_result()`                   | CVA biplot + prediction grid              |
| 2     | 7    | `bl_find_boundary()`                  | Global counterfactual search              |
| 2     | 8    | `plot(bl_bnd)`                        | Variable importance, distance to boundary |
| 2     | 9    | `bl_surrogate()`                      | Spatial surrogate model                   |
| 3     | 10   | `bl_predict()` + `bl_select_target()` | Select target patient                     |
| 3     | 12   | `set_filters()`                       | Encode clinical actionability constraints |
| 3     | 13   | `bl_find_local_cf()`                  | Nearest constrained counterfactual        |
| 3     | 14   | `plot(bl_local)`                      | Local biplot                              |
| 3     | 15   | `bl_shapley()`                        | Shapley attribution                       |
| 3     | 16   | `bl_find_sparse_cf()`                 | Sparse counterfactual                     |
| —     | 17   | Unconstrained search                  | Baseline comparison                       |
| —     | 18   | External patient                      | New observation not in training data      |

The key finding from the Shapley analysis is which clinical variables
**Support** moving the patient from the diabetic region toward the
non-diabetic decision boundary, quantified as probability contributions.
The sparse counterfactual identifies the minimum set of changes that
would flip the model's prediction, expressed in interpretable clinical
units.
